Multiset-trie data structure

### 2019-06-19 tex/tex/s_experiments.tex
Small addition to Introduction. Related work.

% anouncement of the experiments
In the next sections we will present the behavior of the multiset-trie data 
structure depending on the selected parameters as well as the comparative 
benchmark of the multiset-trie against B+ tree implementation of inverted index. 
% outline of the experiment section
We start with experiments that are performed on an artificially generated data in 
order to give a general picture of the multiset-trie performance. In the 
Experiment~\hyperref[s:exp1]{1} a special case of the multiset-trie is considered. 
Only sets are allowed to be stored in the data structure, i.e. the maximal allowed 
multiplicity is set to 1. The performance is measured with respect to the density 
of the multiset-trie.
%
The Experiment~\hyperref[s:exp2]{2} is an extension of the previous one. Here, 
we also measure the performance of the multiset-trie depending on its density. 
The difference is that the allowed multiplicity of an element is raised, i.e. 
the data structure is populated with multisets. 
%
Summarizing the tests of performance depending on the density we present the 
Experiment~\hyperref[s:exp3]{3}. It shows a non linearity of the performance 
with respect to the density of the multiset-trie.
%
The next experiment on the multiset-trie uses the real world data. In 
Experiment~\hyperref[s:exp4]{4} the influence of the mapping $\phi$ is studied. 
The input data is obtained by mapping of the real words from English dictionary 
to the set of consecutive integers using the function $\phi.$ The experiment 
shows that the performance of the multiset-trie is noticeably influenced by 
different mappings $\phi.$ It also shows the usability of the multiset-trie in terms 
of real data demonstrating the high performance of search queries.
%
After all the experiments we present an empirical comparison of multiset-trie 
data structure with B+ tree based inverted index. We use inverted index 
to store and retrieve multisets in the same way as it is described in the paper by 
Helmer and Moerkotte~\cite{Helmer2003} for sets. In the comparison we use 
three types of queries exact, submultiset and supermultiset retrieval.

@@

\subsection{Experiment 5} \label{s:exp5}
In this experiment, we demonstrate the performance of multiset-trie data structure compared to the inverted index based on the B+ tree. Both data structures are implemented in the programming language \CC, providing in this way an experimental setup for a fair comparison \cite{akulich2019mstrie}.

((Brief description of the implementation of the inverted files?))

>>>>

% anouncement of the experiments
In the next sections we will present the behavior of the multiset-trie data 
structure depending on the selected parameters as well as the comparative 
benchmark of the multiset-trie against B-tree implementation of inverted index. 
% outline of the experiment section
We start with experiments that are performed on an artificially generated data in 
order to give a general picture of the multiset-trie performance. In the 
Experiment~\hyperref[s:exp1]{1} a special case of the multiset-trie is considered. 
Only sets are allowed to be stored in the data structure, i.e. the maximal allowed 
multiplicity is set to 1. The performance is measured with respect to the density 
of the multiset-trie.
%
The Experiment~\hyperref[s:exp2]{2} is an extension of the previous one. Here, 
we also measure the performance of the multiset-trie depending on its density. 
The difference is that the allowed multiplicity of an element is raised, i.e. 
the data structure is populated with multisets. 
%
Summarizing the tests of performance depending on the density we present the 
Experiment~\hyperref[s:exp3]{3}. It shows a non linearity of the performance 
with respect to the density of the multiset-trie.
%
The next experiment on the multiset-trie uses the real world data. In 
Experiment~\hyperref[s:exp4]{4} the influence of the mapping $\phi$ is studied. 
The input data is obtained by mapping of the real words from English dictionary 
to the set of consecutive integers using the function $\phi.$ The experiment 
shows that the performance of the multiset-trie is noticeably influenced by 
different mappings $\phi.$ It also shows the usability of the multiset-trie in terms 
of real data demonstrating the high performance of search queries.
%
After all the experiments we present an empirical comparison of multiset-trie 
data structure with B-tree based inverted index. We use inverted index 
to store and retrieve multisets in the same way as it is described in the paper by 
Helmer and Moerkotte~\cite{Helmer2003} for sets. In the comparison we use 
three types of queries exact, submultiset and supermultiset retrieval.

@@

\subsection{Experiment 5} \label{s:exp5}
In this experiment, we demonstrate the performance of multiset-trie data structure compared to the inverted index based on the B-tree. Both data structures are implemented in the programming language \CC, providing in this way an experimental setup for a fair comparison \cite{akulich2019mstrie}.

%((Brief description of the implementation of the inverted files?))
The inverted index is implemented using an idea from~\cite{Helmer2003}. An inverted index structure consist of two parts: a dictionary and postings. In our case a dictionary is implemented as an in-memory B-tree where keys are all distinct values from a domain represented by a set $\Sigma.$ The postings are represented by lists of multisets that contain a particular element from $\Sigma.$ Each list item in postings contains a cardinality of a multiset, which speeds up the containment queries.

### 2019-06-19 tex/tex/s_intro.tex
Small addition to Introduction. Related work.

The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} or, the B+ tree~\cite{Helmer2003}. These solutions provide a key-value look-up for elements of a multiset implying that containment operations are still element-wise. (? Is it possible to state more precisely the drawbacks of these solutions? ?)

>>>>

The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} or, the B+ tree~\cite{Helmer2003}. These solutions provide a key-value look-up, where a key is an element of a multiset and a value is a corresponding multiset implying that containment operations require additional processing of partial results such as intersection or union.

### 2019-06-19 tex/tex/s_relwork.tex
algorithms adjusted and get-all algorithms added
Small addition to Introduction. Related work.

\section{Related work} \label{c:relwork}
[Will be rewritten]
%
% Set theory, multiest theory, bag-of-words model
\subsection{Multiset}
Multiset is a widely used data structure in different areas of mathematics, physics 
and computer science \cite{singh2007overview}. The theory of multisets is based 
entirely on the theory of sets. However, classical mathematics does not deal with 
multisets directly. Instead, one can define a multiset to be a \emph{family} of sets 
or the functions on ordered pairs, where the members of a pair are an element and its 
multiplicity. This means that mathematically the concepts of set such as cardinality, 
set-containment operation, power set, equivalence classes and others are well defined 
for multisets in terms of sets \cite{blizard1988multiset}.

The concept of a multiset can also be referred to the \emph{bag-of-words model}. 
This model takes its origin from a linguistic context studied by Harris~\cite{harris1954distributional}. 
According to the bag-of-words model, text can be represented as a bag (multiset) 
of words, where an element is a word and the number of its occurrences in the 
text is multiplicity. A bag of words does not keep track of grammar and ordering 
of words.

% Information retrieval
\subsection{Information retrieval}
\emph{Information retrieval (IR)} refers to a problem of finding material of an 
unstructured nature that satisfies an information need~\cite{manning2008introduction}. 
Usually, one is searching for a specific documents in a significantly large text 
documents database. The size of a database makes the search a time consuming 
operation. In order to resolve the issue IR systems pre-process data and create 
indexes for future use in search operation.

The bag-of-words model is widely used in IR. In particular, such a representation of 
text documents is used in database indexes when a full text search of a database 
is required. The full-text search problem refers to indexing techniques for full-text 
databases. The most efficient index nowadays uses the concept of an inverted index 
\cite{zobel1992efficient}.

The proposed data structure multiset-trie can be used as an alternative implementation 
of the search structure of an inverted index. It represents words as multisets and 
stores them into data structure. The query processing is achieved using boolean 
retrieval model \cite{manning2008introduction} and multiset containment operations. 
Multiset containment operations of the multiset-trie implement the nearest neighbor 
search queries which retrieve not only exact but also the most relevant results 
to a user.

\subsection{Generalized search tree}
The properties and operations of the multiset-trie makes it a competitor to the 
most efficient implementation of a search tree the \emph{Generalized Search Tree 
(GiST)}~\cite{broder2006indexing,hellerstein1995generalized,kornacker1999high} 
that is used in inverted index. GiST is a very flexible data structure that can be 
customized in order to behave like B+-tree, R-tree or RD-tree. It also provides 
support for an extensible set of queries and data types that B+-tree, R-tree or RD-tree 
do not support originally. GiST supports all the basic search tree operations such as 
insert, delete and search, and in addition provides such extensions as the 
nearest-neighbor search and multiset containment operations. The extensions 
provided by GiST are native in the multiset-trie. The multiset-trie also has a fixed 
height while GiST is a self-balanced tree and has to use additional methods in 
order to preserve its balance.

% Set-trie
\subsection{Set-trie}
The multiset containment queries are well studied in the area of relational 
databases. The queries are well-defined in the context of relational 
algebra~\cite{Lamperti2000msetsinds}. This problem was previously studied 
for a restricted case of queries. In particular, the storage and fast retrieval of sets 
were previously accomplished in data structure \emph{set-trie} proposed by 
Savnik~\cite{savnik2013index}.

The set-trie data structure is based on a trie data structure. It supports set 
containment operations such as retrieval of the nearest sub- and supersets and 
retrieval of all sub- and supersets from the data structure.

The data structure multiset-trie adapts the properties of the set-trie implementing 
the functions \textsc{submsetExistence}, \textsc{supermsetExistence}, 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets} together with the basic 
tree functions such as \textsc{insert}, \textsc{delete} and \textsc{search}. 
Moreover, multiset-trie extends the abilities of the set-trie allowing to store and 
retrieve multisets. The downside of such an extension is that multiset-trie no 
longer supports path compression that was obtained in set-trie. However, the 
design of multiset-trie provides a constant worst case time complexity of search 
function independently of user input.

>>>>

\section{Related work} \label{c:relwork}

% intro to related work

Our work is related to storage of multisets and multiset containment queries. In this section we summarize previous work done 
for storage and retrieval of sets and multisets as well as work on (multi)set containment queries. Also, we discuss the applications 
of multiset-trie with respect to already existing data structures.

% Construction of multiset-trie

\subsection{Set-trie}
% set-trie: 
% - data structure that multiset-trie generalizes
% - stores a set of sets
% - supports set containment operations

The multiset-trie is closely related to the set-trie data structure proposed by Savnik~\cite{savnik2013index}. 
It is a trie-based data structure that is adapted to efficient storage and retrieval of sets. In particular, 
it stores a set of sets and supports set containment operations such as retrieval of the nearest sub- 
and supersets as well as retrieval of all sub- and supersets from the data structure.

The data structure multiset-trie generalizes the set-trie allowing to store a set of multisets. When the multiset-trie 
is restricted to store a set of sets the underlying data structure becomes a simple binary tree. Moreover, all the 
operations of the set-trie are also supported by the multiset-trie. The generalization is, of course, comes with a small penalty 
in performance if we compare the multiset-trie with the set-trie. The downside of such a generalization is that multiset-trie 
no longer supporting path compression that was obtained in set-trie. However, the design of multiset-trie provides a storage 
of multisets as well as constant worst-case time complexity of search operation independently of user input.

\subsection{Inverted file}
% inverted index:
% - performance study of set containment queries (sequential signature files, signature trees, extendible signature hashing, inverted files)
% - performance comparison of multiset-trie to inverted index
% - containment query issues of inverted index

The performance study~\cite{Helmer2003} shows that the inverted file is the most efficient data structure for set containment queries 
among data structures: sequential signature files, signature trees, extendible signature hashing and inverted files. 
In our work we replicate the experiment with inverted file and compare the performance of set containment queries between inverted file and multiset-trie data structures. We user the same methodology as in related paper in order to implement containment queries for inverted file.
The results of this comparison can be found the Experiments section~\ref{s:exp5}.

% Application areas of multiset-trie

\subsection{Set containment joins}
[WIP]
% set containment joins:
% - review the paper: Set containment join revisited\cite{bouros2016set}
% - multiset containment join

\subsection{Multisets in relational databases}
[WIP]
% multisets in RDBMS and ORDBMS:
% - set-valued attributes
% - multiset-valued attributes
% - underlying datastructures for storage of set- and multiset-valued attributes

\subsection{Application of multiset-trie}
[WIP]
% application of multiset-trie
% - indexing of multisets
% - containment queries
% - multiset containment join
% - information retrieval

%
% Set theory, multiest theory, bag-of-words model
%\subsection{Multiset}
%Multiset is a widely used data structure in different areas of mathematics, physics 
%and computer science \cite{singh2007overview}. The theory of multisets is based 
%entirely on the theory of sets. However, classical mathematics does not deal with 
%multisets directly. Instead, one can define a multiset to be a \emph{family} of sets 
%or the functions on ordered pairs, where the members of a pair are an element and its 
%multiplicity. This means that mathematically the concepts of set such as cardinality, 
%set-containment operation, power set, equivalence classes and others are well defined 
%for multisets in terms of sets \cite{blizard1988multiset}. 
%
%The concept of a multiset can also be referred to the \emph{bag-of-words model}. 
%This model takes its origin from a linguistic context studied by Harris~\cite{harris1954distributional}. 
%According to the bag-of-words model, text can be represented as a bag (multiset) 
%of words, where an element is a word and the number of its occurrences in the 
%text is multiplicity. A bag of words does not keep track of grammar and ordering 
%of words.

% Information retrieval
%\subsection{Information retrieval}
%\emph{Information retrieval (IR)} refers to a problem of finding material of an 
%unstructured nature that satisfies an information need~\cite{manning2008introduction}. 
%Usually, one is searching for a specific documents in a significantly large text 
%documents database. The size of a database makes the search a time consuming 
%operation. In order to resolve the issue IR systems pre-process data and create 
%indexes for future use in search operation. 
%
%The bag-of-words model is widely used in IR. In particular, such a representation of 
%text documents is used in database indexes when a full text search of a database 
%is required. The full-text search problem refers to indexing techniques for full-text 
%databases. The most efficient index nowadays uses the concept of an inverted index 
%\cite{zobel1992efficient}. 
%
%The proposed data structure multiset-trie can be used as an alternative implementation 
%of the search structure of an inverted index. It represents words as multisets and 
%stores them into data structure. The query processing is achieved using boolean 
%retrieval model \cite{manning2008introduction} and multiset containment operations. 
%Multiset containment operations of the multiset-trie implement the nearest neighbor 
%search queries which retrieve not only exact but also the most relevant results 
%to a user.

%\subsection{Generalized search tree}
%The properties and operations of the multiset-trie makes it a competitor to the 
%most efficient implementation of a search tree the \emph{Generalized Search Tree 
%(GiST)}~\cite{broder2006indexing,hellerstein1995generalized,kornacker1999high} 
%that is used in inverted index. GiST is a very flexible data structure that can be 
%customized in order to behave like B+-tree, R-tree or RD-tree. It also provides 
%support for an extensible set of queries and data types that B+-tree, R-tree or RD-tree 
%do not support originally. GiST supports all the basic search tree operations such as 
%insert, delete and search, and in addition provides such extensions as the 
%nearest-neighbor search and multiset containment operations. The extensions 
%provided by GiST are native in the multiset-trie. The multiset-trie also has a fixed 
%height while GiST is a self-balanced tree and has to use additional methods in 
%order to preserve its balance.
